---
title: "TSAR Project Assignment Part1"
subtitle: "Exploratory Data Analysis with ggplot2"
author: "Philipp Altenbach, Taoufik Brinis, Ronny Grieder, Ryan Kreis"
date : today
#warning: false
format:
  html:
    theme: zephyr
    code-fold: true
    page-layout: full
#format: pdf #Needs tinytex or similar
editor: visual #Visual editor by default
execute:
  echo: true  # Show code in output by default. If false no code only output.
---

```{r setup, include=FALSE}
# install the required packages if needed
# need to be commented because myLCdata file must not always be created again
# use only myLCdata <- fread("Data/myLCdata.csv")

# if (!require("data.table")) install.packages("data.table") if (!require("dplyr")) install.packages("dplyr") \
# load them 
library(data.table)
library(dplyr)
library(here)
library(skimr)
library(ggplot2)
library(plotly)
library(GGally)
# 
# LCdata <- fread("Data/LCdata.csv")
# 
# set.seed(5)
# 
# myLCdata <- LCdata %>% sample(5) %>% slice_sample(prop = .5)
# 
# fwrite(myLCdata, file = "Data/myLCdata.csv")
```

```{r include=FALSE}
data_path <- file.path(here(), "Data", "myLCdata.csv")
print(here())
myLCdata <- fread(data_path)
```

# Task 1. Business Understanding

Getting a first impression by using:

-   `glimpes(myLCdata)`

```{r}
# Usage of glimpes() to get a first overview over our 5 features
glimpse(myLCdata)

```

-   `str(myLCdata)`

```{r}
# Usage of str() to get further insights into our dataset
str(myLCdata)
```

# Task 2. Data Understanding Based on Summary Statistics

Review of the main statistical metrics of our underlying attributes.

```{r}
summary(myLCdata)
skim(myLCdata)
```

## Interpretation of `int_rate`

The `int_rate` attribute does not show any anomalies. The spread between min and max is relatively tight which indicates a rather symmetrical bell shape that follows a normal distribution. If the distribution is really close to normal, it indicates that most borrowers are categorized into risk levels evenly. Therefore, the dataset does not seam to have a strong bias toward either high or low-risk borrowers.

## Interpretation of `mths_since_last_delinq`

The attribute `mths_since_last_delinq` shows 22816 missing values. However, put in the domain context a missing value likely indicates that the borrower has **never defaulted** rather than it being a data error. In practice, lenders might interpret these missing values as **a positive sign of creditworthiness**.

## Interpretation of `revol_util`

The attribute `revol_util` shows 33 missing values. However, this values is a rather small proportion of the dataset and may not be of significant issue. There is **moderate variability (Standard Deviation: 23.87)**, meaning some borrowers utilize much more or much less of their available credit. The median is very close to the mean, indicating a relatively **balanced** distribution. The min value of 0 suggest that some borrowers do not use any of the resolving credit at all. When analyzing the numbers under the domain knowledge further implication regarding the financial & credit risk of costumers can be derived as followed:

-   **Borrowers with low utilization (\<30%)** are generally seen as **low risk** in credit scoring models, as they manage debt well.

-   **Borrowers with moderate utilization (\~30-70%)** may still be financially stable but are using a significant portion of their available credit.

-   **Borrowers with high utilization (\>70%)** are at **higher risk** since lenders see high credit usage as a potential sign of financial stress.

-   **Utilization \>100%** suggests **over-the-limit borrowing**, which could indicate poor financial management or reporting errors

## Interpretation of `tot_cur_bal`

The standard deviation is quite high, which indicates significant variability in current balance values. This suggests the presence of outliers, likely individuals with exceptionally high balances, skewing the distribution. Moreover, there are 3558 missing values, indicating that there might be actual missing data.

## Interpretation of `delinq_2yrs`

If a large proportion of values are **zero**, since the median and 1st to 3rd quertile are 0. Additionally, the mean is close to zero, indicating that few borrowers have had delinquencies of plus 30 days in the past two years, which may impact predictive models if not handled properly. It is questionable if this bias does reflect real-world. This bias might lead to bad generalizability since the model might fail to detect risky customers. On the other hand, given the domain context, the data might suggest that delinquency of borrowers is often below 30 day, meaning that outstanding payments are redeemed promptly.

# Task 3. Univariate Exploratory Data Analysis (EDA) with ggplot2

Our underlying dataset contains the following data types:

```{r}
cat(paste(names(myLCdata), "➡️", sapply(myLCdata, class)), sep = "\n")
```

Based on the respective data type of each attribute, the following diagram types were chosen.

## Diagrams for `int_rate`

In task 2 we claimed that the int_rate attribute tends to show a normal distribution. Let's now use some suitable graphs to investigate and proof this claim.

The following Histogram with density plot, as well as the Q-Q Plot aim to visualize the existence of a normal distribution of the attribute int_rate:

```{r}
# Histogram with density plot
ggplot(myLCdata, aes(x = int_rate)) +
  geom_histogram(aes(y= after_stat(density)), bins=30, fill="darkblue", alpha=0.5) +
  geom_density(color="red", linewidth=1) +
  ggtitle("Histogram & Density Plot of int_rate") +
  xlab("Interest Rate") +
  ylab("Density")

# Q-Q Plot
#| label: fig-Q-Q
qqnorm(myLCdata$int_rate, main = "Normal Q-Q Plot of int_rate")
qqline(myLCdata$int_rate, col = "red", lwd = 2)
```

Based on the histogram with density plot the int_rate attribute appears to almost follow a normal distribution. Additionally, many of the points in the Q-Q plot follow the red line, indicating that the distribution is close to normal. Consequently, the assumption and interpretations made in task 2 seem to be true.

## Diagrams for `mths_since_last_delinq`

In task 2 we mentioned the absence of 22816 values. Nonetheless, as stated in task 2 and under the aspect of the domain context, those values can be interpreted **as a positive sign of creditworthiness.** The following bar chart aims to visualize this behavior:

```{r}
myLCdata %>%
  mutate(missing = ifelse(is.na(mths_since_last_delinq), "Missing", "Present")) %>%
  ggplot(aes(x = missing, fill = missing)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  ggtitle("Missing vs. Present Values in mths_since_last_delinq") +
  xlab("Status") +
  ylab("Count")
```

## Diagrams for `revol_util`

In the previous task we indicated that there is moderate variability within the attribute. To reveal if the attribute is evenly distributed, indicating that the dataset contains an evenly distributed amount of borrowers with low to high revolving dept utilization, the following boxplot and histogram were created.

```{r}
# Boxplot for revol_util
boxplot_revol_util <- ggplot(myLCdata, aes(y = revol_util)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", fill = "steelblue") +  
  labs(title = "Boxplot of Revolving Credit Utilization",
       x = "",
       y = "Revolving Credit Utilization (%)") +
  theme_minimal()

# Convert to interactive plot
ggplotly(boxplot_revol_util) 

ggplot(myLCdata, aes(x = revol_util)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  ggtitle("Distribution of Revolving Credit Utilization") +
  xlab("Revolving Credit Utilization (%)") +
  ylab("Frequency")
```

The variability in `revol_util` appears **moderate**, with most values concentrated around the mean but a few significant outliers above 100% utilization.

## Diagrams for `tot_cur_bal`

The boxplot shows a **wide spread of values**, with a significant number of **outliers**. This could indicate high-net-worth borrowers or individuals with large outstanding credits. Additionally, the presence of outliers suggests that the data is **highly variable**, meaning some borrowers carry significantly larger balances than others. In the context of credit risk, higher total current balance could therefore mean higher debt level, which could correlate with financial stress.

```{r}
# Boxplot for tot_cur_bal
boxplot_tot_cur_bal <- ggplot(myLCdata, aes(y = tot_cur_bal)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", fill = "steelblue") +  
  labs(title = "Boxplot of the current balance",
       x = "",
       y = "Current Balance") +
  theme_minimal()
# Convert to interactive plot
ggplotly(boxplot_tot_cur_bal)
```

## Diagrams for `delinq_2yrs` 

The following histogram shows a sharp **peak at zero** confirming that most borrowers have not had delinquencies longer than 30 days within the last two years. This undermines the previously expressed **data bias,** as the imbalance suggest that a predictive model might struggle to generalize to riskier customers.

```{r}

test <- ggplot(myLCdata, aes(x = delinq_2yrs)) +
  geom_histogram(bins = 10, fill = "steelblue", alpha = 0.5) +
  ggtitle("Histogram of Delinquencies (30+ Days Past Due)") +
  xlab("Number of Delinquencies (Past 2 Years)") +
  ylab("Frequency") +
  theme_minimal()

ggplotly(test)

```

# Task 4. Bivariate EDA with ggplot2

Using ggpairs to visualize the relationship between all attributes in the dataset.
```{r}
ggpairs(myLCdata)
```
##Next we selected the following two types of 2-dimensional subplots that ggpairs() is showing us.

First, we selected the scatter plot of 'int_rate' and 'revol_util' to visualize the relationship between the interest rate and the revolving credit utilization. The scatter plot shows a **positive correlation** between the two attributes, indicating that borrowers with higher credit utilization tend to have higher interest rates. This makes sense, as higher credit utilization is often associated with higher risk and therefore higher interest rates.
```{r}
library(ggplot2)
library(viridis)

ggplot(myLCdata, aes(x = revol_util, y = int_rate)) +
  geom_point(alpha = 0.5, color = "#1f78b4", size = 2) +  # Points first
  geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"), se = TRUE, color = "black", linewidth = 1.5) +  # Thick, black trend line
  labs(
    title = "Interest Rate vs. Revolving Credit Utilization",
    subtitle = "Higher credit utilization may lead to higher interest rates.",
    x = "Revolving Utilization (%)",
    y = "Interest Rate (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_blank()
  )

```
Next we chose...
```{r}

```