---
title: "P3-5"
subtitle: "Times Series Forecasting"
author: "Philipp Altenbach, Taoufik Brinis, Ronny Grieder, Ryan Kreis"
date: today
date-format: long
format:
  html:
    theme: zephyr
    code-fold: show
    page-layout: full
  # pdf:
  #   code-overflow: wrap        # Prevent code overflow
  #   number-sections: true      # Section Numbers
  #   code-line-numbers: true    # Show line numbers in code
  #   fig-width: 7               # Adjust figure width
  #   fig-height: 5              # Adjust figure height
  #   fig-pos: "H"               # Keep figures in place
  #   fig-align: center          # Align figures at the center
  #   tbl-cap-location: top      # Table captions at the top
  #   tbl-colwidths: auto        # Allow tables to auto-adjust width
  #   geometry: a4paper, margin=1.0in # A4 page with slightly reduced margins
execute:
  echo: true                   # Show code in output
  warning: false               # Suppress warnings
  message: false               # Suppress messages
editor: visual
---

```{r setup, include=FALSE}

# install required packages if needed
if (!require("fpp3")) install.packages("fpp3")
if (!require("data.table")) install.packages("data.table")
if (!require("dplyr")) install.packages("dplyr")
if (!require("DataExplorer")) install.packages("DataExplorer")
if (!require("tidyr")) install.packages("tidyr")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("plotly")) install.packages("plotly")
if (!require("scales")) install.packages("scales")


# loading required packages
library(fpp3)
library(plotly)

```

```{r Prelimiaries, warning=FALSE}

# Brief inspection of `us_employment`
View(us_employment)

# Loading our underlying timeseries using ID: CEU1021000001
mySeries <- us_employment %>%
  filter(Series_ID == "CEU1021000001")

# Autoplot and first inspection of mySeries
View(mySeries)
autoplot(mySeries, Employed)

# Filter for potential NAs
mySeries %>% filter(is.na(Employed)) %>% head()

# Removing NA
mySeries <- mySeries %>%
  filter(!is.na(Employed))

# Inspection of mySeries after NA removal
View(mySeries)
autoplot(mySeries, Employed)

# Saving series on disk
saveRDS(mySeries, file = "CEU1021000001_series.rds")
```

# 1. Extracting the Training Set

```{r extracting training set}
# Load the saved time series
mySeries <- readRDS("CEU1021000001_series.rds")

# multiplying by 0.8 to get training data
n_total <- nrow(mySeries)
n_train <- floor(0.8 * n_total)

# storing train
train <- mySeries %>% 
  slice(1:n_train) # slice rows from indices 1 to n_train
```

# 2. Training 1st Benchmark Model (Drift Method)

```{r 1st bechnmark model}
# Train model on training data using drift method
benchmark_model_1 <- train %>%
  model(
    drift = RW(Employed ~ drift())
  )

# Inspect fitted values using augment()
fitted_benchmark1 <- benchmark_model_1 %>%
  augment()
```

# 3. Evaluating the Model Fit of the 1st Benchmark Model

## a. Creating a time plot "Actual vs. Fitted"

```{r actual-vs-fitted}
# Time Plot "Actual vs. Fitted"
af_1 <- fitted_benchmark1 %>%
  autoplot(Employed, color = "darkblue") +
  autolayer(fitted_benchmark1, .fitted, color = "red", linetype = "dashed") +
  labs(title = "Actual vs. Fitted Values (Drift Model)",
    y = "Employed",
    x = "Date") +
  theme_minimal()

# Make interactive using plotly
ggplotly(af_1)
```

**Are the fitted values close to the actual values?**

-   Yes, the fitted values are usually fairly close to the actual values, with
    some exceptions (e.g., Oct. 1971 and Apr. 1978).
    
**Are they systematically too high or too low?**

-   The fitted values do not seem to be systematically too high or too low.

**If yes, where does it come from? (Consider how your benchmark model works!)**

-   The previous question was answered with not, so, therefore, the drift 
    method seems to be a good fit.

## b. Creating a scatter plot "Actual vs. Fitted":

```{r scatter plot Actual vs. Fitted}
sp_1 <- fitted_benchmark1 %>%
  ggplot(aes(x = .fitted, y = Employed)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, linetype = 2, color = "red") +
  labs(title = "Actual vs. Fitted (Drift Model)",
       x = "Fitted Values",
       y = "Actual Values") +
  theme_minimal()

ggplotly(sp_1)

```

**Are the points close to the identity line?**

-   Yes, with some exceptions, the points generally cluster around the identity 
    line, which indicates a reasonably good overall fit.

**Are they systematically too high or too low?**

-   There are no strong systematic deviations, but some points diverge and might form outliers.

**If yes, where does it come from? (Relate your answer to what you saw in the time plot.)**

-   Outliers might occur because the drift model captures long-term trends but fails to adjust for short-term fluctuations.

## c. Performing residual diagnostics to inspect the model fit:

```{r residual diagnostics for 1st benchmark model}
# Residual diagnostics for the drift model
benchmark_model_1 %>%
  gg_tsresiduals() # Will throw warning because of one NA row.
```

**Are the residuals auto-correlated? How do you decide that based on the plots?**

-   Yes, the residuals show signs of autocorrelation, since several clear spikes in the residual ACF plot exceed the significance bounds.

**Do the residuals have zero mean? How do you decide that based on the plots?**

-   Yes, the histogram of residuals is centered around zero, and the time plot of residuals fluctuates symmetrically around the zero line

**What do these results tell you about your model?**

-   The model captures the overall trend but overlooks short-term dependencies, which suggests untapped predictive information within the data and the need for an even more sophisticated approach.

## d. Double-checking the results from (c).

```{r ljung-box-and-residual-mean}
# Ljung-Box test for residual autocorrelation
# Using lag = 24
fitted_benchmark1 %>%
  features(.resid, ljung_box, lag = 24, dof = 1)

# Residual mean (applied mean() to .resid column)
mean(fitted_benchmark1$.resid, na.rm = TRUE)
```

**Does the Ljung-Box test result support your conclusions from (c)?**

-   Yes, the Ljung-Box test confirms significant autocorrelation in the residuals 
(p-value = 0.000007979894), which supports the earlier observation from the ACF plot.

**Does the residual mean result support your conclusions from (c)?**

-   The residual mean is effectively zero (–2.97e-14), which confirms the model 
is unbiased (property 2 of residual diagnostics), but this alone does not confirm 
the absence of autocorrelation.

# 4. Evaluating the Point Forecast Accuracy of the 1st Benchmark Model

```{r forecast-drift}
forecast_length <- n_total - n_train # Remainder from TS after training data extracted
# Forecast for 149 months ahead
fc_benchmark1 <- benchmark_model_1 %>%
  forecast(h = forecast_length)

# Time plot of "Actual vs. Forecast"
fc_benchmark1 %>%
  autoplot(mySeries) +
  labs(title = "Actual vs Forecast (Drift Model)",
       y = "Employed",
       x = "Date") +
  theme_minimal()
```

```{r}
# Calculating accuracy on full data
fc_benchmark1 %>%
  accuracy(mySeries)
```

## Interpretation of accuracy metrics

The root mean squared error (RMSE) of approximately 105 is scale-dependent and should be interpreted in the context of the time series. Based on the time plot, this error appears to be roughly in the range of seasonal fluctuations, which is acceptable for a benchmark model but too large for forecasting within-season employment changes. The mean absolute percentage error (MAPE) of 9.3% compares each forecast error to the level of the series and is appropriate here due to the presence of a strong trend and moderate cyclical behavior. However, while a 9% average error is reasonable for a simple benchmark, it remains inadequate for precise, policy-relevant forecasting.

# 5. Evaluating the Point Forecast Uncertainty of the 1st Benchmark Model

```{r residual-diagnostics-uncertainty}
# Residual diagnostics for evaluating forecast uncertainty
benchmark_model_1 %>%
  gg_tsresiduals()
```

## Are the residuals homoscedastic? (That is, do they have constant variance?)

-   No, the residuals are not entirely homoscedastic, since several sharp spikes 
around 1980 indicate periods of temporary variance inflation.

## Are the residuals normally distributed?

-   Roughly yes, the histogram of residuals appears bell-shaped and centered 
around zero, though there is some asymmetry and evidence of heavy tails in both directions.

```{r prediction-intervals}
# Display prediction intervals (80% and 95%) for forecast
fc_benchmark1 %>%
  hilo()
```

## Does it make sense to include these prediction intervals in your model evaluation? Why/why not?

-   No, it does not make sense to include these prediction intervals in the model
evaluation, since the presence of autocorrelation and heteroscedasticity in the 
residuals undermines the reliability of the forecast uncertainty estimates.


## Training 2nd Benchmark Model (Seasonal Naive Method)
```{r 2nd benchmark model}
# Train model on training data using seasonal naive method
benchmark_model_2 <- train %>%
  model(
    s_naive = SNAIVE(Employed)
  )

# Inspect fitted values using augment()
fitted_benchmark2 <- benchmark_model_2 %>%
  augment()
```

# 3. Evaluating the Model Fit of the 1st Benchmark Model

## a. Creating a time plot "Actual vs. Fitted"
```{r}
# Time Plot "Actual vs. Fitted"
af_2 <- fitted_benchmark2 %>%
  autoplot(Employed, color = "darkblue") +
  autolayer(fitted_benchmark2, .fitted, color = "red", linetype = "dashed") +
  labs(title = "Actual vs. Fitted Values (Seasonal Naive Model)",
    y = "Employed",
    x = "Date") +
  theme_minimal()

# Make interactive using plotly
ggplotly(af_2)
```
**Are the fitted values close to the actual values?**
-   In some instances, especially when predictable cycles occur, the fitted values 
are somewhat close to the actual values, however, during fluctuations or unexpected c
hanges, the model fails to predict accurately and is often relatively far off.
    
**Are they systematically too high or too low?**
-   We were not able to identify a consistent directional bias (too high or too low), 
but there appears to be a systematic lag in the model's predictions.

**If yes, where does it come from? (Consider how your benchmark model works!)**
-   The fitted values are very similar or identical to the values from the same 
month in the previous year, which explains the consistent shift or lag in the predictions.

## b. Creating a scatter plot "Actual vs. Fitted":
