---
title: "TSAR Project Assignment Part1"
subtitle: "Exploratory Data Analysis with ggplot2"
author: "Philipp Altenbach, Taoufik Brinis, Ronny Grieder, Ryan Kreis"
date : today
#warning: false
format:
  html:
    theme: zephyr
    code-fold: true
    page-layout: full
#format: pdf #Needs tinytex or similar
editor: visual #Visual editor by default
execute:
  echo: true  # Show code in output by default. If false no code only output.
---

```{r setup, include=FALSE}
# install the required packages if needed
# need to be commented because myLCdata file must not always be created again
# use only myLCdata <- fread("Data/myLCdata.csv")

# install required packages if needed
if (!require("data.table")) install.packages("data.table")
if (!require("dplyr")) install.packages("dplyr")
if (!require("here")) install.packages("here")
if (!require("skimr")) install.packages("skimr")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("plotly")) install.packages("plotly")
if (!require("GGally")) install.packages("GGally")

# loading required packages
library(data.table)
library(dplyr)
library(here)
library(skimr)
library(ggplot2)
library(plotly)
library(GGally)

# Important: Set WD to Part_1_EDA
# TSAR-ID: 5

# LCdata <- fread("Data/LCdata.csv")
# set.seed(5)
# myLCdata <- LCdata %>% sample(5) %>% slice_sample(prop = .5)
# fwrite(myLCdata, file = "Data/myLCdata.csv")

myLCdata <- fread("Data/myLCdata.csv")
```

```{r dynamic path setup, include=FALSE}
# data_path <- file.path(here(), "Data", "myLCdata.csv")
# print(here())
# myLCdata <- fread(data_path)
```

# Task 1. Business Understanding

Using `dplyr::glimpse()` and `str()` to get a first impression of the data set.

## `dplyr glimpse()`

```{r glimpse()}
# Usage of glimpes() to get a first overview over our 5 features
dplyr::glimpse(myLCdata)
```

## `str()`

```{r str()}
# Usage of str() to get further insights into our dataset
str(myLCdata)
```

## Understanding our Sampled Features

Most our sampled features of the data set are self-explanatory when consulting the data dictionary. However, at first, the feature `revol_util` can be somewhat challenging to interpret. It represents the percentage of available revolving credit a borrower is using, such as on a credit card. A higher value means more credit is being utilized, while a lower value indicates more available credit. Furthermore, using the sampling technique described above, \`myLCdata' contains only numerical attributes.

# Task 2. Data Understanding Based on Summary Statistics

Using `summary()` and `skimr::skim()` to review the main statistical metrics of our attributes.

## `summary()` 
```{r summary()}
summary(myLCdata)
```

## `skimr::skim()`
```{r skimr::skim()}
skimr::skim(myLCdata)
```

## Interpretation of `int_rate`

The `int_rate` attribute does not show any anomalies. The spread between min and max is relatively tight, indicating a rather symmetrical bell shape that follows a normal distribution. If the distribution is close to normal, it suggests that most borrowers are categorized into risk levels evenly. Therefore, the dataset does not seem to have a strong bias toward either high or low-risk borrowers.

## Interpretation of `mths_since_last_delinq`

The attribute `mths_since_last_delinq` shows 22816 missing values. However, in the domain context, a missing value likely indicates that the borrower has **never defaulted** rather than it being a data error. In practice, lenders might interpret these missing values as **a positive sign of creditworthiness**.

The skewness of `mths_since_last_delinq` indicates a slight **right-skewed** distribution. This means that while most borrowers have relatively recent delinquencies, there are a few with long durations since their last delinquency.

## Interpretation of `revol_util`

The attribute `revol_util` shows 33 missing values, which is a relatively small proportion of the dataset and may not be a significant issue. There is **moderate variability (Standard Deviation: 23.87)**, meaning some borrowers utilize much more or much less of their available credit. The median is very close to the mean, indicating a relatively **balanced** distribution. The minimum value of 0 suggests that some borrowers do not use any revolving credit at all. Additionally, the maximum **value (148.00)** compared to the 75th percentile (73.90) suggests **potential outliers**, which could influence variability interpretation and measures.

From a financial risk perspective:

-   **Borrowers with low utilization (\<30%)** are generally seen as **low risk** in credit scoring models, as they manage debt well.
-   **Borrowers with moderate utilization (\~30-70%)** may still be financially stable but use a significant portion of their available credit.
-   **Borrowers with high utilization (\>70%)** are at **higher risk**, as lenders see high credit usage as a potential sign of financial stress.
-   **Utilization \>100%** suggests **over-the-limit borrowing**, which could indicate poor financial management or reporting errors.

## Interpretation of `tot_cur_bal`

The standard deviation is quite high (`157,273.25`), indicating substantial variability in current balance values. This strongly suggests the presence of **outliers**, which might be individuals with exceptionally high balances, who are skewing the distribution. Furthermore, there are **3,558 missing values (8% of the data).** If the dataset was intended wo be further processesd, imputation or exclusion might be required depending on the analytical approach.

The distribution of `tot_cur_bal` is **highly skewed to the right**, which can be seen by the large discrepancy between the mean (`139,427.49`) and median (`78,813.00`). Additionally, when applying the IQR-Rule, there are **approximately 1,400 outliers**, primarily high-balance borrowers, which could impact predictive modeling if not properly addressed through transformations.

## Interpretation of `delinq_2yrs`

A large proportion of values are **zero**, as the median and interquartile range all indicate no delinquencies. The mean is also close to zero, suggesting that most borrowers have had **no** 30+ days past-due incidences of delinquency in the past two years. However, the presence of a high number of "solid" borrowers could potentially introduce biases in predictive models if not handled properly, potentially reducing generalizability.

The dataset reveals over **8000 outliers** for `delinq_2yrs`. Given the quartile values, most of these outliers are likely **borrowers with exactly 1 delinquency**, which is a key insight. The presence of only a few cases with higher values suggests that **delinquencies of more than 1 occurrence are rare** in this dataset.

# Task 3. Univariate Exploratory Data Analysis (EDA) with ggplot2

Our underlying data set contains the following data types:

```{r data types}
cat(paste(names(myLCdata), "➡️", sapply(myLCdata, class)), sep = "\n")
```

Based on the respective data type of each attribute, the following diagram types were chosen.

## Diagrams for `int_rate`

In task 2 we claimed that the int_rate attribute tends to show a normal distribution. Let's now use some suitable graphs to investigate and proof this claim.

The following Histogram with a density plot and the theoretical normal density function (orange dashed line), as well as the Q-Q Plot (used to compare the feature's distribution) aim to visualize the existence of a normal distribution of the attribute int_rate:

```{r plots int_rate}
# Calculate mean and standard deviation of int_rate
mean_int_rate <- mean(myLCdata$int_rate, na.rm = TRUE)
sd_int_rate <- sd(myLCdata$int_rate, na.rm = TRUE)

# Histogram with density plot and normal curve
ggplot(myLCdata, aes(x = int_rate)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "steelblue", alpha = 0.5) +
  geom_density(color = "red", linewidth = 1) +  # Empirical density curve
  stat_function(fun = dnorm, args = list(mean = mean_int_rate, sd = sd_int_rate), 
                color = "orange", linewidth = 1.2, linetype = "dashed") +  # Normal curve
  ggtitle("Figure 1: Histogram & Density Plot of int_rate with Normal Curve") +
  xlab("Interest Rate") +
  ylab("Density") +
  theme_minimal()

# Q-Q Plot
#| label: fig-Q-Q
qqnorm(myLCdata$int_rate, main = "Figure 2: Normal Q-Q Plot of int_rate", pch = 16, col = "steelblue")
qqline(myLCdata$int_rate, col = "red", lwd = 2)
```

Based on the histogram with the density plot, the int_rate attribute shows a roughly bell-shaped distribution but exhibits slight right skewness. While many points in the Q-Q plot follow the red line, deviations at the tails suggest that the distribution is not perfectly normal. However, the assumption of approximate normality made in Task 2 still holds reasonably well for most of the data, with some caution needed for outliers.

## Diagrams for `mths_since_last_delinq`

In Task 2, we mentioned the absence of 22,816 values. Nonetheless, as stated in Task 2 and under the aspect of the domain context, those values can be interpreted as a **positive sign of creditworthiness**. The following bar chart visualizes this missing vs. present ratio. Additionally, the boxplot illustrates the **distribution of `mths_since_last_delinq`**, considering **only the present (non-missing) values** to focus on the actual reported delinquencies.

```{r plots mths_since_last_delinq}
# Bar chart for missing vs. present values
myLCdata %>%
  mutate(missing = ifelse(is.na(mths_since_last_delinq), "Missing", "Present")) %>%
  ggplot(aes(x = missing, fill = missing)) +
  geom_bar() +
  scale_fill_manual(values = c("Missing" = "red", "Present" = "steelblue")) +
  ggtitle("Figure 4: Missing vs. Present Values in mths_since_last_delinq") +
  xlab("Status") +
  ylab("Count")

# Boxplot for only present values
ggplot(myLCdata %>% filter(!is.na(mths_since_last_delinq)), aes(y = mths_since_last_delinq)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", fill = "steelblue") +
  ggtitle("Figure 5: Boxplot of mths_since_last_delinq (Only Present Values)") +
  ylab("Months Since Last Delinquency") +
  theme_minimal()


```

## Diagrams for `revol_util`

In the previous task we indicated that there is moderate variability within the attribute. To reveal if the attribute is evenly distributed, indicating that the dataset contains an evenly distributed amount of borrowers with low to high revolving dept utilization, the following boxplot and histogram were created. Note: This will produce a warning for 33 observations, which correspond to the missing values described above.

```{r plots revol_util, warning=FALSE}
# warning=FALSE --> suppress warnings in rendered version
# Boxplot for revol_util
boxplot_revol_util <- ggplot(myLCdata, aes(y = revol_util)) +
  geom_boxplot(alpha = 0.7, fill = "steelblue") +  
  labs(title = "Figure 6: Boxplot of Revolving Credit Utilization",
       x = "",
       y = "Revolving Credit Utilization (%)") +
  theme_minimal()

# Convert to interactive plot
ggplotly(boxplot_revol_util) 

# Histogram for revol_util
ggplot(myLCdata, aes(x = revol_util)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  ggtitle("Figure 7: Distribution of Revolving Credit Utilization") +
  xlab("Revolving Credit Utilization (%)") +
  ylab("Frequency")
```

The variability in `revol_util` appears **moderate**, with most values concentrated around the mean but a few significant outliers above 100% utilization.

## Diagrams for `tot_cur_bal`

The boxplot shows a **wide spread of values**, with a significant number of **outliers**. This could indicate high-net-worth borrowers or individuals with large outstanding credits. Additionally, the presence of outliers suggests that the data is **highly variable**, meaning some borrowers carry significantly larger balances than others. In the context of credit risk, higher total current balance could therefore mean higher debt level, which could correlate with financial stress.

```{r plots tot_cur_bal, warning=FALSE}
# Boxplot for tot_cur_bal
boxplot_tot_cur_bal <- ggplot(myLCdata, aes(y = tot_cur_bal)) +
  geom_boxplot(alpha = 0.7, fill = "steelblue") +  
  labs(title = "Figure 8: Boxplot of the current balance",
       x = "",
       y = "Current Balance") +
  theme_minimal()
# Convert to interactive plot
ggplotly(boxplot_tot_cur_bal)
```

## Diagrams for `delinq_2yrs`

The following histogram shows a sharp **peak at zero**, confirming that most borrowers have not experienced delinquencies of more than 30 days in the past two years. This challenges the previously mentioned **data bias**, as the imbalance suggests that a predictive model may struggle to generalize to riskier customers. In addition, a box plot was created for this variable to examine outliers, that is, borrowers who had more than zero delinquencies over 30 days in the past two years.


```{r plots delinq_2yrs, warning=FALSE}
hist_delinq_2yrs <- ggplot(myLCdata, aes(x = delinq_2yrs)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.5) +
  ggtitle("Figure 9: Histogram of Delinquencies (30+ Days Past Due)") +
  xlab("Number of Delinquencies (Past 2 Years)") +
  ylab("Frequency") +
  theme_minimal()

ggplotly(hist_delinq_2yrs)

boxplot_delinq_2yrs <- ggplot(myLCdata, aes(y = delinq_2yrs)) +
  geom_boxplot(alpha = 0.7, fill = "steelblue") +
  ggtitle("Figure 10: Boxplot of Delinquencies (30+ Days Past Due)") +
  ylab("Number of Delinquencies (Past 2 Years)") +
  theme_minimal()

ggplotly(boxplot_delinq_2yrs)

```

# Task 4. Bivariate EDA with ggplot2

Using ggpairs to visualize the relationship between all attributes in the dataset.

## Create Pairs Plot
```{r ggpairs, warning=FALSE}
ggpairs(myLCdata, columns = c("int_rate", "revol_util", "mths_since_last_delinq", "delinq_2yrs")) +
  ggtitle("Figure 11: Selected Variable Relationships")
```

## Analyzing Insightful Subplots

Next we selected the following two types of 2-dimensional subplots that ggpairs() is showing us.

### Interest Rate `int_rate` vs. Revolving Credit Utilization `revol_util`

One notable relationship is between **interest rate** `int_rate` and **revolving credit utilization** `revol_util`, which shows a moderate **positive correlation of 0.271**. This suggests that borrowers who utilize a larger portion of their available credit tend to be charged higher interest rates. Lenders may perceive high credit utilization as a sign of greater financial dependency, which increases the risk associated with the borrower and justifies a higher loan cost.

### Scatterplot with Regression Line (`revol_util` vs. `int_rate`)

```{r scatter revol_util vs. int_rate, warning=FALSE, message=FALSE}
# Scatterplot: Interest Rate vs. Revolving Credit Utilization
ggplot(myLCdata, aes(x = revol_util, y = int_rate)) +
  geom_point(alpha = 0.4, color = "steelblue") +  
  geom_smooth(color = "red") +  
  ggtitle("Figure 12: Interest Rate vs. Revolving Credit Utilization") +
  xlab("Revolving Credit Utilization (%)") +
  ylab("Interest Rate (%)") +
  theme_minimal()
```

This scatterplot illustrates the **relationship between revolving credit utilization `revol_util` and interest rate `int_rate`**.

### Months Since Last Delinquency `mths_since_last_delinq` vs. Number of Delinquencies in Past 2 Years `delinq_2yrs`

Another significant finding is the **negative correlation (-0.561)** between `mths_since_last_delinq` and `delinq_2yrs`. This indicates that borrowers with more recent delinquencies tend to have a higher number of past delinquencies within the last two years. The pattern suggests that borrowers who have defaulted recently are more likely to have a history of repeated delinquencies, which plays into their classification as high-risk individuals.

### Scatterplot with Regression Line (`mths_since_last_delinq` vs. `delinq_2yrs`)

```{r scatter mths_since_last_delinq vs. delinq_2yrs, warning=FALSE, message=FALSE}
ggplot(myLCdata, aes(x = mths_since_last_delinq, y = delinq_2yrs)) +
  geom_point(alpha = 0.4, color = "steelblue") +  
  geom_smooth(color = "red") +  
  ggtitle("Figure 13: Months Since Last Delinquency vs. Delinquencies in the Past 2 Years") +
  xlab("Months Since Last Delinquency") +
  ylab("Number of Delinquencies (Past 2 Years)") +
  theme_minimal()
```

Similarly, this scatterplot visualizes the strong negative correlation (-0.561) between `mths_since_last_delinq` and `delinq_2yrs`. Borrowers who have had a delinquency recently tend to have more delinquencies in the past two years.

## Enhanced Scatterplot using Revolving Credit Utilization `revol_util` as a Third Variable

We added `revol_util` as a third variable to the `mths_since_last_delinq` vs. `delinq_2yrs` plot to explore its impact on delinquencies. Credit utilization is a key financial indicator—borrowers with high `revol_util` may struggle more with repeated delinquencies. 

This visualization helps reveal whether **high credit utilization correlates with frequent delinquencies**, which provides deeper insight into borrower risk profiles beyond just delinquency timing.


```{r scatter third variable, warning=FALSE, message=FALSE}
ggplot(myLCdata, aes(x = mths_since_last_delinq, y = delinq_2yrs, color = revol_util)) +
  geom_point(alpha = 0.6) +  
  geom_smooth(color = "red") +  
  scale_color_gradient(low = "steelblue", high = "green", name = "Revol. Credit Utilization (%)") +  
  ggtitle("Figure 15: Delinquencies vs. Months Since Last Delinquency (Revol. Credit Utilization)") +
  xlab("Months Since Last Delinquency") +
  ylab("Number of Delinquencies in Past 2 Years") +
  theme_minimal()
```
